---
title: "Take-home Exercise 2: Application of Spatial and Spatio-temporal Analysis Methods to Discover the Distribution of Dengue Fever in Tainan City, Taiwan"
execute:
  eval: true
  echo: true
  warning: false
toc-depth: 2
code-annotations: hover
---

# 1.0 Introduction

## 1.1. Overview - Setting the Scene

[Dengue Hemorrhagic Fever](https://www.cdc.gov/dengue/resources/denguedhf-information-for-health-care-practitioners_2009.pdf) (in short dengue fever) is one of the most widespread mosquito-borne diseases in the most tropical and subtropical regions. It is an acute disease caused by dengue virus infection which is transmitted by female Aedes aegypti and Aedes albopictus mosquitoes. In 2015, Taiwan had recorded the most severe dengue fever outbreak with more than 43,000 dengue cases and 228 deaths. Since then, the annual reported dengue fever cases were maintained at the level of not more than 200 cases. However, in 2023, Taiwan recorded 26703 dengue fever cases.

## 1.2 Objectives

In this study, we are interested to discover:

-   If the distribution of dengue fever outbreak at Tainan City, Taiwan are independent from space and space and time.

-   If the outbreak is indeed spatial and spatio-temporal dependent, then, you would like to detect where are the clusters and outliers, and the emerging hot spot/cold spot areas.

## 1.3 Getting Started

In this take-home exercise, we will be using the following packages.

```{r}
pacman::p_load(tidyverse, tmap, sf, sp, sfdep)
```

# 2.0 Data Acquisition

We will be using **2 data sets** in this exercise:

| Data                | Format         | Description                                                    | Source                                                                                               |
|-------------|-------------|-----------------|------------------------------|
| TAIWAN_VILLAGE_2020 | ESRI Shapefile | A geospatial data of village boundary of Taiwan                | [data.gov.tw](https://data.gov.tw/en/datasets/130549)                                                |
| Dengue_Daily.csv    | csv            | An aspatial data of reported dengue cases in Taiwan since 1998 | [Taiwan CDC Open Data Portal](https://data.cdc.gov.tw/en/dataset/dengue-daily-determined-cases-1998) |

Create a new folder labelled `Take-home_Ex02` and place all the files into this folder.

# 3.0 Importing Geospatial and Aspatial Data

## 3.1 Importing Geospatial data - TAIWAN_VILLAGE_2020

The TAIWAN_VILLAGE_2020 dataset was acquired in ESRI shapefile format (.shp). To utilise this dataset in the R-environment, we need to import it as an sf object using the st_read() function from the sf package. This function is used to read the shapefile containing the administrative boundaries of Tainan City and returns an sf object named `tainan_sf`.

```{r}
tainan_sf <- st_read(dsn = "data/geospatial", 
                layer = "TAINAN_VILLAGE")
```

## 3.2 Importing Aspatial data - Dengue_Daily

The Dengue_Daily dataset is available in csv format (.csv) and was obtained from the Taiwan CDC Open Data Portal. Like the previous dataset, it needs to be imported into the R environment for use. However, since this dataset is aspatial and in csv format, a different method is required for reading it. We will utilize the read_csv() function to import the csv cdataset and store the object in a tibble data frame named `dengue`.

```{r}
dengue <- read_csv("data/aspatial/Dengue_Daily.csv")
head(dengue)
```

# 4.0 Data Wrangling

## 4.1 Preparing a study area layer with specific counties of Tainan City, Taiwan

**Task: Confining study area layer to D01, D02, D04, D06, D07, D08, D32 and D39 counties**

In this Take-home Exercise, we are interested in narrowing our focus to specific counties within Tainan City, specifically **D01, D02, D04, D06, D07, D08, D32, and D39**. To prepare a study area layer focusing on these specific counties, we can do the following:

```{r}
counties <- c('D01', 'D02', 'D04', 'D06', 'D07', 'D08', 'D32', 'D39')
tainan_counties_sf<- tainan_sf %>% filter(TOWNID %in% counties)
```

::: callout-note
## Notes

The c() function is used to combine the specified counties into a vector (a one dimensional array) named counties. Next, we will filter the Tainan City spatial data frame (`tainan_sf`) based on the `TOWNID` column, selecting only those entries that match the counties of interest listed in the counties vector. This refined dataset, named `tainan_counties_sf`, will serve as our study area layer for further analysis or visualization tasks.
:::

Note: Check that tainan_counties_sf only contains polygon features.

## 4.2 Preparing a dengue fever layer with specific counties of Tainan City, Taiwan

**Tasks:**

1.  **Extracting dengue fever cases within epidemiology week 31-50, 2023**
2.  **Confining dengue fever layer to D01, D02, D04, D06, D07, D08, D32 and D39 counties**

### 4.2.1 **Extracting dengue fever cases within epidemiology week 31-50, 2023**

```{r}
dengue_2023 <- dengue %>%
  select(發病日,
         最小統計區中心點X,
         最小統計區中心點Y) %>%
  mutate(EPIWEEK = epiweek(發病日)) %>%
  filter(year(dengue$發病日) == 2023 & 最小統計區中心點X != "None" & 最小統計區中心點Y != "None")
```

Visualizing the distribution of dengue fever cases across the epidemiology weeks

```{r}
ggplot(dengue_2023, aes(x = EPIWEEK)) +
  geom_histogram(binwidth = 1, color = "grey") +
  labs(x = "EPIWEEK", y = "count") +
  ggtitle("Distribution of Dengue Cases in 2023 by Epidemiology weeks") +
  theme_minimal()
```

More than 80% of the reported dengue fever cases occurred in epidemiology week 31-50, 2023.

```{r}
dengue_2023_epiweeks_31_50 <- dengue_2023 %>%
  filter(between(EPIWEEK, 31, 50))

unique(dengue_2023_epiweeks_31_50$EPIWEEK)
```

### 4.2.2 Converting aspatial data (converting to sf tibble data.frame)

```{r}
dengue_2023_epiweeks_31_50_sf <- st_as_sf(dengue_2023_epiweeks_31_50,
                                          coords = c("最小統計區中心點X", "最小統計區中心點Y"),
                                          crs = 4326) %>%
  st_transform(crs = 3826)

glimpse(dengue_2023_epiweeks_31_50_sf)
```

Note: Check that dengue_2023_epiweeks_31_50_sf only contains point features.

### 4.2.3 Checking the coordinate systems

::: panel-tabset
## tainan_counties_sf

```{r}
st_crs(tainan_counties_sf)
```

## dengue_2023_epiweeks_31_50_sf

```{r}
st_crs(dengue_2023_epiweeks_31_50_sf)
```
:::

### 4.2.4 Project Transformation - Missing/Inaccurate Coordinate System

tainan_counties_sf is in geographic coordinate systems. We have to change it to projected coordinate system with EPSG 3826.

```{r}
tainan_counties_sf <- st_transform(tainan_counties_sf, st_crs(3826))
st_crs(tainan_counties_sf)
```

Observe that both sfs are in the same coordinate systems.

### 4.2.5 Confining dengue fever layer to D01, D02, D04, D06, D07, D08, D32 and D39 counties

```{r}
dengue_fever <- st_intersection(dengue_2023_epiweeks_31_50_sf, tainan_counties_sf)
```

```{r}
st_geometry(dengue_2023_epiweeks_31_50_sf)
```

```{r}
st_geometry(tainan_counties_sf)
```

```{r}
st_geometry(dengue_fever)
```

```{r}
tm_shape(tainan_counties_sf) +
  tm_polygons("TOWNID") +
tm_shape(dengue_fever) +
  tm_symbols(size = 0.02, col = "black")
```
