---
title: "Take-home Exercise 2: Application of Spatial and Spatio-temporal Analysis Methods to Discover the Distribution of Dengue Fever in Tainan City, Taiwan"
execute:
  eval: true
  echo: true
  warning: false
toc-depth: 2
code-annotations: hover
---

# 1.0 Introduction

## 1.1. Overview - Setting the Scene

[Dengue Hemorrhagic Fever](https://www.cdc.gov/dengue/resources/denguedhf-information-for-health-care-practitioners_2009.pdf) (in short dengue fever) is one of the most widespread mosquito-borne diseases in the most tropical and subtropical regions. It is an acute disease caused by dengue virus infection which is transmitted by female Aedes aegypti and Aedes albopictus mosquitoes. In 2015, Taiwan had recorded the most severe dengue fever outbreak with more than 43,000 dengue cases and 228 deaths. Since then, the annual reported dengue fever cases were maintained at the level of not more than 200 cases. However, in 2023, Taiwan recorded 26703 dengue fever cases.

## 1.2 Objectives

In this study, we are interested to discover:

-   if the distribution of dengue fever outbreak at Tainan City, Taiwan are independent from space and space and time.

-   If the outbreak is indeed spatial and spatio-temporal dependent, then, you would like to detect where are the clusters and outliers, and the emerging hot spot/cold spot areas.

## 1.3 Getting Started

In this take-home exercise, we will be using the following packages.

```{r}
pacman::p_load(tidyverse, tmap, sf, sp)
```

# 2.0 Data Acquisition

We will be using **2 data sets** in this exercise:

| Data                | Format         | Description                                                    | Source                                                                                               |
|-------------|-------------|-----------------|------------------------------|
| TAIWAN_VILLAGE_2020 | ESRI Shapefile | A geospatial data of village boundary of Taiwan                | [data.gov.tw](https://data.gov.tw/en/datasets/130549)                                                |
| Dengue_Daily.csv    | csv            | An aspatial data of reported dengue cases in Taiwan since 1998 | [Taiwan CDC Open Data Portal](https://data.cdc.gov.tw/en/dataset/dengue-daily-determined-cases-1998) |

Create a new folder labelled `Take-home_Ex02` and place all the files into this folder.

# 3.0 Importing Geospatial and Aspatial Data

## 3.1 Importing Geospatial data - TAIWAN_VILLAGE_2020 in shapefile format

The TAIWAN_VILLAGE_2020 dataset was acquired in ESRI shapefile format (.shp). To utilise this dataset in the R-environment, we need to import it as an sf object using the st_read() function from the sf package. This function is used to read the simple features or layers from file or database and returns an sf object.

```{r}
tainan_sf <- st_read(dsn = "data/geospatial", 
                layer = "TAINAN_VILLAGE")
```

```{r}
st_crs(tainan_sf)
```

## 3.2 Importing Aspatial data - Dengue_Daily.csv

The Dengue_Daily dataset is available in CSV format and was obtained from the Taiwan CDC Open Data Portal. Like the previous dataset, it needs to be imported into the R environment for use. However, since this dataset is aspatial and in CSV format, a different method is required for reading it. We will utilize the read_csv() function to import the CSV dataset.

```{r}
dengue <- read_csv("data/aspatial/Dengue_Daily.csv")
```

# 4.0 Data Wrangling

## 4.1 Preparing a study area layer with specific counties of Tainan City, Taiwan

```{r}
counties <- c('D01', 'D02', 'D04', 'D06', 'D07', 'D08', 'D32', 'D39')
tainan_counties_sf<- tainan_sf %>% filter(TOWNID %in% counties)
```

## 4.2 Extracting dengue fever cases with epidemiology week 31-50, 2023

```{r}

```

## 4.3 Preparing a dengue fever layer with specific counties of Tainan City, Taiwan

```{r}

```
