---
title: "Hands-on Exercise 10: Modelling Geographic of Accessibility"
format: html
execute:
  eval: true
  echo: true
  warning: false
date: 03/23/2024
code-annotations: hover
toc-depth: 3
---

# 1.0 Introduction

## 1.1 Getting Started

In this hands-on exercise, we will learn how to **model geographically accessibility**.

## 1.2 Overview

## 1.3 Installing and loading R packages

```{r}
pacman::p_load(tmap, SpatialAcc, sf, 
               ggstatsplot, reshape2,
               tidyverse)
```

# 2.0 Data Acquisition

# 3.0 Geospatial Data Handling

We will be using the *st_read()* from **sf** package to import the data into RStudio.

## 3.1 Importing Geospatial Data

We will import this data into the R environment using the [*st_read()*](https://www.rdocumentation.org/packages/sf/versions/0.7-2/topics/st_read) function of **sf**.

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MP14_SUBZONE_NO_SEA_PL")
```

```{r}
hexagons <- st_read(dsn = "data/geospatial",
                    layer = "hexagons") 
```

```{r}
eldercare <- st_read(dsn = "data/geospatial",
                     layer = "ELDERCARE") 
```

## 3.2 Importing Aspatial Data

The `OD_Matrix.csv` file is in .csv format. We will be import it using *read_csv* function of **readr** package.

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

## 3.3 Data Preparation
