---
title: "Take-home Exercise 3: Prototyping Modules for Geospatial Analytics Shiny Application"
execute:
  eval: true
  echo: true
  warning: false
format:
  html:
    code-fold: false
date: 03/12/2024
toc-depth: 3
code-annotations: hover
---

# **1.0 Loading Packages**

```{r}
pacman::p_load(sf, st, lubridate, tidyverse, tmap, ggplot2, spatstat)
```

# **2.0 Importing Datasets into R Environment**

In this exercise, we will be using the following datasets:

| Data                                                 | Type           | Format   |
|------------------------------------------------------|----------------|----------|
| ➡️**Thailand Road Accident \[2019-2022\]**           | **Aspatial**   | **.csv** |
| ➡️**Thailand Subnational Administrative Boundaries** | **Geospatial** | **.shp** |
| Geofabrik's OpenStreetMap road data for Thailand     | Geospatial     | .shp     |

## **2.1 Importing Geospatial Data**

The **Thailand Subnational Administrative Boundaries** dataset is available in ESRI shapefile format. It comprises the administrative boundaries at various levels within Thailand.

-   Level 0: country

-   **Level 1: province**

-   Level 2: district

-   Level 3: sub-district or tambon

```{r}
thailand_sf <- st_read(dsn = "data/geospatial", 
                layer = "tha_admbnda_adm0_rtsd_20220121")
```

```{r}
thailand_province_sf <- st_read(dsn = "data/geospatial", 
                layer = "tha_admbnda_adm1_rtsd_20220121")
```

There are a total of 77 provinces in Thailand.

```{r}
tmap_mode("plot")

thailand_map <- tm_shape(thailand_sf) + 
  tm_borders(alpha = 0.8)

thailand_province_map <- tm_shape(thailand_province_sf) + 
  tm_polygons("ADM1_EN") + 
  tm_borders(alpha = 0.8) +
  tm_legend(show=FALSE)

tmap_arrange(thailand_map, thailand_province_map, ncol = 2)
```

## **2.2 Importing Aspatial Data**

```{r}
accidents <- read_csv("data/aspatial/thai_road_accident_2019_2022.csv")
```

# **3.0 Data Wrangling**

## **3.1 Preparing a study area layer by provinces**

### **3.1.1 Extract relevant columns**

Firstly, we use the colnames() to see all the column names present in `thailand_province_sf`.

```{r}
colnames(thailand_province_sf)
```

Next, from `thailand_province_sf`, we will select the following relevant columns so that its easier for our analysis later on.

-   ADM1_EN

-   ADM1_PCODE

-   geometry

```{r}
thailand_sf_extracted <- thailand_province_sf %>%
  select(ADM1_EN,
         ADM1_PCODE,
         geometry)
```

### **3.1.2 Check coordinate system**

### **3.1.3 Transform geographic coordinate system (GCS) and a projected coordinate system (PCS)**

### **3.1.4 Create OWIN object**

### **3.1.5 Plot OWIN object**

## **3.2 Preparing a road accident layer**

### **3.2.1 Extract relevant columns**

```{r}
colnames(accidents)
```

Similar to `thailand_province_sf`, we will select the following relevant columns from `accidents`.

-   latitude

-   longtitude

-   incident_datetime

-   province_en

-   presumed_cause

-   weather_condition

-   number_of_fatalities

-   number_of_injuries

```{r}
accidents_extracted <- accidents %>%
  select(latitude,
         longitude,
         incident_datetime,
         province_en,
         presumed_cause,
         weather_condition,
         number_of_fatalities,
         number_of_injuries)
```

```{r}
str(accidents_extracted)
```

```{r}
colSums(is.na(accidents_extracted))
```

```{r}
accidents_extracted <- na.omit(accidents_extracted)
```

```{r}
colSums(is.na(accidents_extracted))
```

### **3.2.2 Converting aspatial data into geospatial data**

```{r}
accidents_extracted <- st_as_sf(accidents_extracted, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 32648)
```

Checking the number of unique provinces in `accidents_extracted`

```{r}
unique(accidents_extracted$province_en)
```

There are a total of 78 provinces? An unknown province is present in the unique list.

Remove unknown province.

```{r}
accidents_extracted <- accidents_extracted %>% 
  filter(province_en != "unknown")
```

### **3.2.3 Check coordinate system**

```{r}

```

### **3.2.4 Create ppp objects using sf method**

```{r}
accidents_extracted_ppp <- as.ppp(accidents_extracted)
summary(accidents_extracted_ppp)
```

### **3.2.5 Check for duplicated point events**

```{r}
any(duplicated(accidents_extracted_ppp))
```

```{r}
sum(multiplicity(accidents_extracted_ppp) > 1)
```

The output shows that there are 15 duplicated point events.

### **3.2.6 Removing duplicated point events using jittering approach**

```{r}
```
